---
title: "Bioinformatics Midterm Exam 2026"
author: "write your name here"
format: html
editor: visual
css: styles.css
---

## Load libraries

you will need "tidyverse", "broom" and "ggbeeswarm" (hint `library()`) if you miss library install it with `install.packages()`

```{r, message=FALSE}
library(tidyverse)
library(broom)
library(ggbeeswarm)
```

## Load dataset

1.  load "mice_weights_and_diet.tsv" data sets and tissue_gene_expression.tsv

```{r, message=FALSE}
mice_weights <- read_tsv("data/mice_weights_and_diet.tsv")

GeneExpression <- read_tsv("data/tissue_gene_expression.tsv")
```

------------------------------------------------------------------------

## Examine Data

2.  Inspect the  mice_weights object with `head()`, `tail()`

```{r}
mice_weights %>% head()
mice_weights %>% tail()
```

3.  What are the dimensions of mice_weights?

```{r}
dim(mice_weights)
```

### Subsetting Data using base R

4.  Select the 2nd column of mice_weights only using square brackets.

```{r}
mice_weights[,2] %>% head()
```

5.  Select the 42nd row of mice_weights using square brackets.

```{r}
mice_weights[42,]
```

6.  Select rows 10 to 25 and columns 1 and 5 using square brackets from mice_weights.

```{r}
mice_weights[10:25,c(1,5)]
```

7.  Select the `sex` column using a dollar sign from mice_weights

```{r}
mice_weights$sex %>% head
```

8.  Find the the minimum and maximum of mice_weights

```{r}
mice_weights %>%
  summarize(min(body_weight), max(body_weight))
```

9.  Use `arrange()` to rearrange the "percent_fat" table so the largest entry is the first in mice_weights

```{r}
mice_weights %>%
  arrange(desc(percent_fat)) %>% head()
```

10. Find the mean and standard deviation of litter size for female sex (hint: use `filter`, `group_by()`, `summarize`) from mice_weights.

```{r}
mice_weights %>%
  filter(sex == "F") %>% 
  group_by(diet) %>%
  summarize(average = mean(litter),
            stdev = sd(litter))
```

11. Find the find average weights and standard deviation for male and female on different diets (hint: use `group_by()`) in mice_weights dataset.

```{r}
mice_weights %>%
  group_by(sex, diet) %>%
  summarize(average = mean(body_weight),
            stdev = sd(body_weight))
```

------------------------------------------------------------------------

## Graphics

12. load `tissue gene expression` if you didn't do at at step 1 and look at it with `head()`

```{r}
# GeneExpression <- read_tsv("data/tissue_gene_expression.tsv")

GeneExpression %>% head()
```

### scater plot

13. from "Gene Expression" dataset make scatter plot of correlation of two randomly selected genes

```{r}
GeneExpression %>% 
  ggplot(aes(SOX12, INF2)) +
  geom_point()
```
#### Scater plot w/ linear fit

14. add linear trend line to you plot and change the plot theme with `theme_bw()` or `theme_classic()` or `theme_minimal()` or any other `theme_...` extension

```{r}
GeneExpression %>% 
  ggplot(aes(SOX12, INF2)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal()
```

15. Modified previous plot by splitting data by tissue of origin when plotting this correlation and switching off standard error `se = F` option

```{r}
GeneExpression %>% 
  ggplot(aes(SOX12, INF2, color = tissue)) +
  geom_point() +
  geom_smooth(method = "lm", se = F) +
  theme_bw()
```

### Box plot

16. Pick one random gene and make Box plot of it with tissues type on x-axis, expression level on y-axis and color (`fill` color) by tissue.

```{r}
GeneExpression %>% 
  ggplot(aes(x = tissue, y=INF2, fill = tissue)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +
  theme_bw()
```

### Density plot

17. change previous plot of your randomly selected gene to density plot and color by tissue type (set `alpha=0.2` in density plot)

```{r}
GeneExpression %>% 
  ggplot(aes(x=INF2, color = tissue, fill=tissue)) +
  geom_density(alpha=0.2) +
  theme_bw()
```

18 to make previous plot more informative facet this plot by tissue use as theme try to use `theme_minimal()`

```{r}
GeneExpression %>% 
  ggplot(aes(x=INF2, color = tissue, fill=tissue)) +
  geom_density(alpha=0.2) +
  facet_wrap(~ tissue) +
  theme_minimal()
```

------------------------------------------------------------------------

## Statistical test

### T.Test

19. Go back to mice_weights dataset and plot *Box plot* of `body_weight` by diet and sex (use color and faceting)
(e.g. x-axis `sex`, y-axis `body_weight` ,facet by `diet`)
(extra if you want to make it prettier add points to your boxplot with, try tu use `geom_quasirandom(alpha= 0.5)`)
```{r}
mice_weights %>%
  ggplot(aes(x= sex, y = body_weight, fill=diet)) +
  geom_boxplot(outliers = FALSE) +
   geom_quasirandom(alpha= 0.5) +
  facet_wrap( ~ diet) +
  theme_bw()
```

20. In mice_weights dataset: Q: Is there effect of diet on weight? (hint use t.test)
**Plesase answer the question in writing! One sentence is enought.**
```{r}
mice_weights %>% t.test(body_weight ~ diet, data = .) %>% tidy()
```

21. In mice_weights dataset Q: is there effect of sex on weight? (use t.test)
**Plesase answer the question in writing! One sentence is enought.**
```{r}
mice_weights %>% t.test(body_weight ~ sex, data = .) %>% tidy()
```

### ANOVA

22. re-plot box plot step 16. Now using ANOVA-test `aov` test if the expression of your gene is different across tissues types Q: Is there a difference of your gene expression across tissue types? Is at least one tissue different in expression then other?
**Plesase answer the question in writing! One sentence is enought.**
```{r}
GeneExpression %>% 
  ggplot(aes(x = tissue, y=INF2, fill = tissue)) +
  geom_boxplot() +
  geom_quasirandom() +
  theme_bw()

GeneExpression %>% select(tissue, INF2) %>% 
  aov(INF2 ~ as.factor(tissue), data = .) %>% tidy()
```

23. Do post-hoc test of 22. If you see your Anova-test to be significant, at least one tissue has different expression. Which one (two or more) is/are different? 
Q1: Which two tissue are the most different in your gene expression analysis, 
Q2: Is there any pair where the difference in expression is not significant (expression is same or comparable)?
**Plesase answer the question in writing! One sentence is enought.**
```{r}
GeneExpression %>% select(tissue, INF2) %>% 
  aov(INF2 ~ as.factor(tissue), data = .) %>% TukeyHSD() %>% tidy()
```
------------------------------------------------------------------------

### Linear modeling

Test linear dependence of two genes that you previously piked in steps 13-15 from Gene Expression dataset.

24. Redo plot you did at step 14:
-   re-plot gene1 against gene2
```{r}
GeneExpression %>% 
  ggplot(aes(SOX12, INF2)) +
  geom_point() +
  geom_smooth(method = "lm", se = F) +
  theme_bw()
```

25. Test in linear model if this two genes are dependent (correlated) or not: 
Q: Is the expression between your two genes lineary dependent (correlated)? (H0: no correlation)
**Plesase answer the question in writing! One sentence is enought.**
```{r}
GeneExpression %>% 
  lm(INF2 ~ SOX12, data = .) %>% 
  summary() %>% 
  tidy()
```


### Bonus

26. Bonus: In `mice weights dataset`; Can you put together step 20 and 21 in one test that will test for diet and sex simultaneously ?
(hint use linear model with covariet `lm( y ~ x1 + x2 )`)
```{r}
model <- lm(body_weight ~ diet + sex, data = mice_weights)
summary(model) %>% tidy()
```

------------------------------------------------------------------------

### Principal Component Analysis (PCA)

Run PCA on tissue expression dataset preferably use `PCAtools` package

If you do not have PCAtools install them with uncomenting and running following code:
```{r}
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install("PCAtools")
```

```{r}
library("PCAtools")
```

To help you reformat data, here is reformating step 
```{r}
data_4_pca <- GeneExpression %>% 
  select(-tissue) %>% 
  column_to_rownames("sample") %>% 
  t() %>%
  as_tibble(rownames="geneID") %>% 
  column_to_rownames("geneID")


metadata_4_pca <- GeneExpression %>%  # your_gene_expression
  select(sample, tissue) %>% 
  column_to_rownames("sample")
```

27. run PCA

```{r}
pca_expression <- pca(data_4_pca, metadata = metadata_4_pca)
```

28 plot *Screenplot* that shows inportance of copmonents (hint use `screeplot()` function).
Please do not plot all components 1:10 is enought! (hint `components = 1:10`) 
```{r}
screeplot(pca_expression, components = 1:10)
```
29 Make Scatter plot of few first PCs (hint use `biplot` and `colby = 'tissue'`,)
```{r}
biplot(pca_expression,
       title = 'PC1 versus PC2',
       x = 'PC1', y = 'PC2',
       colby = 'tissue',
       hline = 0,
       vline = 0,
       lab = NULL,
       legendPosition = 'right',
       axisLabSize = 10, titleLabSize = 14
       )
```

