---
title: "Class week1"
author: "JM"
format: html
editor: visual
css: styles.css
---

## Import libraries

```{r setup, include=FALSE}
library(tidyverse)
# library(conflicted)
```

# Tidy Data Wrangling with `dplyr`

`dplyr` is a grammar of data manipulation, providing a consistent set of rationally named functions to help you solve the most common data manipulation challenges.

## Sub setting Data in Base `R`

### Tables and Examining Data

```{r , classes="scrolling"}
iris
```
### Quick peak into data

`head()` and `tail()`

```{r}
# look at the first 6 rows
head(iris)

# look at the first 10 rows
head(iris, 10)

# look at the last 6 rows
tail(iris)
```

`dim()` prints the numbers of rows and columns in a table

```{r}
# rows, columns
dim(iris)
```

### Subsetting Tables with Base R

Square brackets `[]`

```{r , classes="scrolling"}
iris[1:10,]
```

Dollar sign sub-setting `$`

```{r}
iris$Sepal.Length
```

More advanced sub-setting

```{r, classes="scrolling"}
# drop the first column
iris[,-1]

# select rows 1 to 5 and columns 2 to 4
iris[1:5,2:4]

# list the exact rows and columns you want
iris[c(37, 49, 132, 12, 77), c(1, 3)]
```

## Basic math oparation

### basic algebraic operation

```{r}
5+1 
  
5*5
  
5**3 
5^3

(5+1)^3
```

### values assignment to variable

```{r}
# both = and <- works
a=5
a <- 5

a^3
```

### vector assignment to variable

```{r}
b=c(1,2,3,4,5)
b+1
```

### matrix assignment to variable

```{r}
A <- data.frame(colum1 = c(1,2,3,4,5), column2=c(5,4,3,2,1))

# updatig colum complicated way, we will not do it this way
A[,1] <- A[,1]+5

A
```

### Logicals

These operators are how you compare values to each other. Keep them in mind for subsetting and filtering going forward.

-   equals `==`
-   not equal to `!=`
-   greater than `>`
-   greater than or equal to `>=`
-   less than `<`
-   less than or equal to `<=`
-   and `&`
-   or `|`

```{r}
# Are the numbers equal?
2 == 5

# Is the first number less than the second number?
2 < 5

# Is the first number greater than the second number?
2 > 5

# Are the numbers NOT equal
2 != 5
```

Subset tables with logicals

```{r , classes="scrolling"}
# get all columns and all rows where sepal length is greater than or equal to 5

# bigger_5 <- iris$Sepal.Length >= 5
# iris[bigger_5,]

iris[iris$Sepal.Length >= 5,]

# get all columns and all rows species is NOT viriginica
iris[iris$Species != 'virginica',]
```

### Piping

The pipe operator `%>%` pipes output from one function to the next. You can use this to chain together as many command/functions as desired. For a simple example, instead of calling `head()` on flights, we could have piped flights to `head()`

```{r}
## all these are equivalent expression

head(iris)

iris |> head()
iris %>% head()
```

You can chain as many pipes as you want together and you can also pipe into `ggplot()`

```{r , classes="scrolling"}

iris_filter <- iris %>% filter(Species == 'setosa')

iris_filter %>% select(Sepal.Length, Sepal.Width)

iris %>% filter(Species == 'setosa') %>% select(Sepal.Length, Sepal.Width)
```

```{r , classes="scrolling"}
 iris %>% 
  filter(Species == 'setosa') %>% 
  select(Sepal.Length, Sepal.Width) %>% 
  filter(Sepal.Length > 5 & Sepal.Width > 2)
```

### Assignment

The assignment operator, `->` saves the current output as an object (aka a data structure) to whatever name you pick. You can save data of any format this way.

##### right hand assigment, by Google Manual bad practice, it exist but do not use it

```{r}
# iris %>% filter(Species == 'setosa') %>% select(Sepal.Length, Sepal.Width) -> new_iris
# 
# new_iris
```

##### left hand assigment, use this one

```{r , classes="scrolling"}
new_iris <- iris %>% filter(Species == 'setosa') %>% select(Sepal.Length, Sepal.Width)

new_iris
```

##### remove object

```{r}
rm(new_iris)
```

<br>

## Data Manipulation with dplyr

### Setup

Install packages

```{r}
# uncomment the packages and install
# install.packages('tidyverse')
# install.packages('conflicted')
```

### dplyr functions

#### `arrange()`

The `arrange()` function allows you to sort the data by some criteria, usually another column in the data.

```{r , classes="scrolling"}
### Before arranging the data, let's remember what it looks like
iris

### arrange() takes, at a minimum, a single column as an argument. It will sort
### the table by that column in order from least to greatest
iris %>%
  arrange(Petal.Width)

### You can sort the data by as many columns as you like, although the changes
### in the order are smaller with each additional column
iris %>%
  arrange(Petal.Width, Petal.Length, Sepal.Length)

### What if you want to sort from largest to smallest? You can use the helper
### function desc() that changes the sort order from largest to smallest.
iris %>%
  arrange(desc(Petal.Width))
```

### `select()`

The `select()` function allows you to pick only the columns you want

```{r , classes="scrolling"}
### For the select() function, you just give it the names of the columns you 
### want. For more complicated selection criteria, you probably need to use 
### select_if() or select_at() which are beyond the scope of this presentation.
iris %>%
  select(Species, Sepal.Length, Sepal.Width)

select_example <- iris %>%
  select(Species:Sepal.Width)

select_example

iris %>%
  select(species = Species, Sepal.Length, Sepal.Width)
```

### `filter()`

The function `filter()` allows you select a subset of the data according to some criteria. This could be by matching a variable like `sample == 's42'` or could be base on some numberic value like `Sepal.Width < 3 | Sepal.Width > 4`

```{r , classes="scrolling"}
### filter takes that name of the column you want to filter on and some 
### logical condition. Here, we'll get only the rows of the table for samples
### with Sepal.Lengths greater than 4.5 cms.
iris %>%
  filter(Sepal.Length > 5.5, Species == 'setosa')

iris %>%
  filter(Sepal.Length > 5.5 | Species == 'setosa')

### You can filter by as many things as you want
iris %>%
  filter(Sepal.Length > 4.5, Species %in% c('virginica', 'versicolor'), 
         Sepal.Width < 3, Petal.Length > 5, Petal.Width > 1.5)
```

### `distinct()`

The function `distinct()` gives you all unique values of a given variable.

```{r , classes="scrolling"}
### distinct() takes the name of the column you want the distinct values of
iris %>%
  distinct(Sepal.Width)

### You can give it as many columns as you want although it becomes less and
### less useful as you add columns
iris %>%
  distinct(Species, Sepal.Width)
```

### `mutate()`

The function `mutate()` adds an additional column onto the table.

```{r}
A <- data.frame(column1 = c(1,2,3,4,5), column2=c(5,4,3,2,1))

A <- A %>% mutate(column1 = column1 + 1)

A
```

##### this is how we would do it with out mutate

```{r}
A <- data.frame(column1 = c(1,2,3,4,5), column2=c(5,4,3,2,1))
A[,1] <- A[,1]+5

A
```

#### `mutate` together with `condition`if_else\`

```{r , classes="scrolling"}
### mutate() needs the name of the new column equal to whatever will go in the
### new column. Frequently that will be a logical statement based on other
### data in the table, like in this example where we call any iris with a sepal
### length less than 5.8 cm small and all the other irises large

iris %>%
  mutate(flower_size = if_else(Sepal.Length < 5.8, 'small', 'large'))

### You can make more than one column in the same mutate call. If you supply a
### single value, it will be used for all rows (recycled).
iris %>%
  mutate(flower_size = if_else(Sepal.Length < 5.8, 'small', 'large'),
         second_column = 'tidyverse_webinar')

iris %>%
  mutate(flower_size = case_when(Sepal.Length < 3 ~ 'small',
                                 Sepal.Length >= 3 & Sepal.Length < 6 ~ 'medium',
                                 .default = 'large'))
```

### `summarize()`

Lets you reduce the data down to some summary statistics or other summary function

```{r}
### summarize() works similarly to mutate(), but instead of adding an 
### additional column with an entry for all rows, it will reduce to the number
### summaries you asked for. Like mutate(), summarize() needs the name of the
### summary column equal to the calculation you're making. Here, we'll make the
### summary column "average" which is equal to the mean of the petal lengths
iris %>%
  summarize(average = mean(Petal.Length))

### For most tidyverse functions, either the British or American spellings can
### be used. Frequently the British spelling comes up by default because Hadley
### Wickham is from New Zealand.
iris %>%
  summarise(average = mean(Petal.Length))

### summarize() is very powerful when used with group_by(). group_by() causes 
### all functions later in the pipeline to be applied over the specified group
iris %>%
  group_by(Species) %>%
  summarize(average = mean(Petal.Length))

### Don't forget to ungroup() because grouping will mess up your code later
iris %>%
  group_by(Species) %>%
  summarize(average = mean(Petal.Length)) %>%
  ungroup()

### You can calculate more than one summary at a time as long as they (for the
### moment) have a single thing as their output
iris %>%
  group_by(Species) %>%
  summarize(avg_pl = mean(Petal.Length),
            avg_pw = mean(Petal.Width),
            sd_pl = sd(Petal.Length),
            sd_pw = sd(Petal.Width)) %>%
  ungroup()
```

#### `count()`

The function `count()` is like a special case of `summarize()`, but all it does is count the number of distinct things in the column(s).

```{r}
### count() takes the name of the column you want
iris %>%
  count(Species)

### It's particularly powerful when combined with group_by(), like summarize() is
iris %>%
  mutate(flower_size = ifelse(Sepal.Length < 5.8, 'small', 'large')) %>%
  group_by(flower_size) %>%
  count(Species)

### You can also count() multiple things at once
iris %>%
  mutate(flower_size = ifelse(Sepal.Length < 5.8, 'small', 'large')) %>%
  count(Species, flower_size)
```
