## Install if you do not have them yet
uncoment the part you need 
```{r}
# BiocManager::install('tidyverse')
# BiocManager::install('PCAtools')
# BiocManager::install('edgeR')
# #
# BiocManager::install('pheatmap')
# #
# # # Make sure you have devtools installed
# install.packages("devtools")
# # # Then install using devtools::install_github
# devtools::install_github("coriell-research/coriell")
```

## Load Libraries

```{r, message=FALSE}
library('tidyverse')
library('PCAtools')
library('edgeR')
library('coriell')
```

## Load data

```{r}
getwd()
```

## Load Data

##### fill up the rest of file name `...` with file that coresponds to first letter of your first name

```{r, message=FALSE}
tcga_counts <- read_tsv("data/tcga_prad_counts_ ...... ")

metadata <- read_tsv("data/tcga_prad_metadata_ ....... ")
```

### Visualy inspect how dat alooks like

```{r}
tcga_counts %>% head()
```

```{r}
metadata %>% head()
```

# Here are the name of your samples
```{r}
colnames(tcga_counts)
```

### Make version with matrix with column/row names

```{r}
tcga_counts2 <- tcga_counts %>% column_to_rownames("gene_ID")
tcga_counts2 %>% head()
```

```{r}
metadata2 <- metadata %>% column_to_rownames("sample_ID")
metadata2 %>% head()
```

## Make edgeR object for differential gene expression analysis

```{r}
# Initialize the edgeR object
dge <- DGEList(counts = tcga_counts2,
               group = metadata2$type,
               genes = rownames(tcga_counts2))

#  normalization steps independent from our experimental design
dge <- calcNormFactors(dge)
```


## PCA
#### Create logcounts (log counts per milion) for PCA

```{r}
# logcounts not filtered by expression for QC plots
logcounts <- cpm(dge, log = TRUE)

logcounts[1:5,1:6]
```

#### run PCA with PCAtools
```{r}
pca <- pca(logcounts, metadata = dge$samples, center = TRUE, scale = TRUE) #, removeVar = 0) # , removeVar = 0.1
```

### plot PCA results

note: when you saving result picture/table into `results` folder you have to create it first in proper place
either you create it yourself at your computer or type "1" (Yes) at your RStudio console when asked.

```{r}
screeplot(pca, components = 1:12) #, axisLabSize = 18, titleLabSize = 22)

ggsave("results/screeplot_your_XXXgroup.png")
ggsave("results/screeplot_your_XXXgroup.pdf")

# note: when you saving into `results` folder you have to create it first in proper place
```

```{r}
biplot(pca, 
       colby = "group",
       colkey = c(tumor = 'dodgerblue2', adj_norm = 'gold2'),
       # showLoadings = TRUE,
       legendPosition = 'right',
       # encircle = TRUE,
       lab = NULL
       )

ggsave("results/pca_plot_your_XXXgroup.png")
```

## Run diff gene experssion analysis

### Build model

```{r}
# Compare genotypes, controlling for gender
design <- model.matrix( ~ type, data = metadata)

# Determine which genes have enough counts to keep around
keep <- filterByExpr(dge, design = design)
```

```{r}
y <- dge[keep, keep.lib.sizes = FALSE]
y <- normLibSizes(y)
```

### check how many uninformative genes you filter out by looking to final dimention of data
```{r}
dge %>% dim()
y %>% dim()
```

### Check data before Normalization
```{r}
logcounts %>% 
  as_tibble(rownames = "geneID") %>% 
  pivot_longer(cols = -1, names_to = "sample_ID", values_to = "counts") %>% 
  ggplot(aes(sample_ID, counts)) +
  geom_boxplot(outliers = F, fill = "red3", fatten = 4, lwd=0.1) + #, color="gray80"
  theme_classic() +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank())

ggsave("results/normalization_original_data_XXXgroup.png")
```

### Check sample distribution after Normalization
```{r}
# logcounts not filtered by expression for QC plots
logcounts_y <- cpm(y, log = TRUE)

logcounts_y %>% 
  as_tibble(rownames = "geneID") %>% 
  pivot_longer(cols = -1, names_to = "sample_ID", values_to = "counts") %>% 
  ggplot(aes(sample_ID, counts)) +
  geom_boxplot(outliers = F, fill = "red3", fatten = 4, lwd=0.1) + #, color="gray80"
  theme_classic() +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank())

ggsave("results/normalization_normalized_data_XXXgroup.png")
```

### Fit model to data

```{r}
# Fit model
fit <- glmQLFit(y, design, robust = TRUE)
# plotQLDisp(fit)
```

### Get diff expression by running `glmQLFTest`
```{r}
# Test for differences between ANY group -- like a one-way ANOVA
y <- glmQLFTest(fit, coef = 2)
results <- topTags(y, n = Inf, sort.by = "none")$table
```

### Save the results
note: when you saving into `results` folder you have to create it first in proper place
```{r}
results
results %>% write_csv("results/results_XXXgroup.csv")
#write_csv(results, "results/results_table.csv")
```

### PLot DE results

#### MA plot
note: when you saving into `results` folder you have to create it first in proper place
```{r}
# Create and MA plot of the results
m <- plot_md(results, lfc = 1 ,fdr = 0.001)
m

ggsave("results/MA_plot_XXXgroup.png")
```

#### Volcano plot with couts of sig chnged genes
```{r}
plot_volcano(results, x = "logFC", y = "FDR", fdr = 0.001, lfc = 1,
  #lab = "gene_name", label_sig = TRUE, annotate_counts = FALSE
  ) +
  labs(
    title = "PRAD: tumor vs normal",
    #subtitle = "",
    x = "logFC",
    y = "-log10(PValue)",
    caption = "results for FC>2 and FDR < 0.01"
  ) +
  theme_coriell()

ggsave("results/volcanoplot.plot1_XXXgroup.png")
```

#### Volcano plot with highlight of few most changed genes
```{r}
plot_volcano(results, x = "logFC", y = "FDR", fdr = 0.001, lfc = 1,
  lab = "genes", label_sig = TRUE, annotate_counts = FALSE) +
  labs(
    title = "PRAD: tumor vs normal",
    #subtitle = "",
    x = "logFC",
    y = "-log10(PValue)",
    caption = "results for FC>2 and FDR < 0.01"
  ) +
  theme_coriell()

ggsave("results/volcanoplot_plot2_XXXgroup.png")
```


### Heatmap

#### pull only significan genes

```{r}
siq_genes <- results %>% filter(abs(logFC) > 1 & FDR < 0.001) %>% rownames()
```

##### Check default heatmap
```{r}
# pheatmap::pheatmap(logcounts[siq_genes,])
quickmap(logcounts_y[siq_genes,])
```

#### Nicer version of version of heatmap
note: when you saving into `results` folder you have to create it first in proper place

```{r}
col_df <- metadata %>% 
  select(sample_ID, type) %>% 
  column_to_rownames("sample_ID")

ann_cols <- list(type = c("tumor" = "deepskyblue1", "adj_norm" = "goldenrod1"))

qm <- quickmap(logcounts[siq_genes,],
         annotation_col = col_df,
         annotation_colors = ann_cols,
         show_colnames = FALSE,
         main = "PRAD Tumor vs Normal"
  )

ggsave("results/heatmap_XXXgroup.png", qm)
```
