---
title: "Exam Demo"
author: "JM"
format: html
editor: visual
---

## Demo document

This is the blueprint of code you can use for work with your data for

## Load libraries

```{r}
library(tidyverse)
library(broom)
```

### to oriend where are you in your computer

```{r}
getwd()
```

## Load data

path to your data should exist if you have problem to load data look in which directory are you in your computer by running `getwd()`

After loading red text is just message about data no error

```{r}
wine_data <- read_tsv("data/wine_data_set.tsv")
```

## Examine Data

1.  Look at the `head()`

```{r}
head(wine_data)

# or

wine_data %>% head()
```

2.  and the `tail()`

```{r}
tail(wine_data)

# or

wine_data %>% tail()
```

3.  what are the dimensions of iris?

```{r}
dim(wine_data)
```

### Subsetting Data using base R

4.  Select the first column only using square brackets.

```{r}
wine_data[,1] %>% head()
```

5.  Select the 42nd row using square brackets.

```{r}
wine_data[42,]
```

6.  Select rows 35 to 40 and columns 1 and 3 using square brackets.

```{r}
wine_data[c(35:40), c(1:3)]
```

7.  Select the `Type` column using a dollar sign

```{r}
wine_data$Type  %>%

  head()
```

------------------------------------------------------------------------

### Combining

We'll be using data from the `coronavirus` package for today's assignment. The `coronavirus` dataset has the following columns:

-   `date` = The date of the summary
-   `province` = The province or state, when applicable
-   `country` = The country or region name
-   `lat` = Latitude point
-   `long` = Longitude point
-   `type` = the type of case (i.e., confirmed, death)
-   `cases` = the number of daily cases (corresponding to the case type)

You need to install and load a package to get the data, so make sure you pay attention to the set up steps and run every chunk.

### Set Up

Install the package with the data

```{r}
# install.packages("coronavirus")
```

Load the library. Libraries are collections of functions (and sometimes data like here) that are add ons to base R.

```{r}
library(coronavirus)
```

------------------------------------------------------------------------

We'll be using data from the `coronavirus` package for today's assignment. The `coronavirus` dataset has the following columns:

-   `date` = The date of the summary
-   `province` = The province or state, when applicable
-   `country` = The country or region name
-   `lat` = Latitude point
-   `long` = Longitude point
-   `type` = the type of case (i.e., confirmed, death)
-   `cases` = the number of daily cases (corresponding to the case type)

You need to install and load a package to get the data, so make sure you pay attention to the set up steps and run every chunk.

Load the data. The `data()` loads data from packages into R.

```{r}
# use the data() function to load the data in the package
data("coronavirus")
```

Look at the data. Run the chunk and it will give you a nicely formatted preview of the data.

```{r}
coronavirus %>% head()
```

### Combining `dplyr` Functions

In this second section, the questions get a little harder. You'll need to give more than one argument to the dplyr functions and chain multiple functions together. Again, use the code chunk below each question to give your answer.

------------------------------------------------------------------------

8.  Find the minimum and maximum numbers of cases.

```{r}
coronavirus %>%
  summarize(min(cases), max(cases))
```

9.  Use `arrange()` to rearrange the coronavirus table so the latest date is first. (By latest I mean the closest data to today, for example April 1 2020-04-01 is closer to today than January first 2020-01-01).

```{r}
coronavirus %>%
  arrange(desc(date)) %>%

  head()
```

10. Find the total number of cases per country

```{r}
coronavirus %>%
  group_by(country) %>%
  summarize(total_cases = sum(cases)) %>%
  ungroup() %>%

  head()
```

11. Find the total number of cases per date

```{r}
coronavirus %>%
  group_by(date) %>%
  summarize(total_cases = sum(cases)) %>%
  ungroup()   %>%

  head()
```

12. Find the number of cases by country and type

```{r}
coronavirus %>%
  group_by(country, type) %>%
  summarize(total_cases = sum(cases)) %>%
  ungroup() %>%

  head()
```

13. Find the country with the greatest number of cases so far

```{r}
coronavirus %>%
  group_by(country) %>%
  summarize(total_cases = sum(cases)) %>%
  ungroup() %>%
  arrange(desc(total_cases)) %>%

  head()
```

14. Find the number of cases per Canadian province.

```{r}
coronavirus %>%
  filter(country == 'Canada') %>%
  group_by(country, province) %>%
  summarize(total_cases = sum(cases))
```

15. Find the mean and standard deviation of the cases by country

```{r}
coronavirus %>%
  group_by(country) %>%
  summarize(average = mean(cases),
            stdev = sd(cases)) %>%

  head()
```

16. Add a column to the `coronavirus` table labelling whether the daily case total is greater than or equal to 100

```{r}
coronavirus %>% 
  mutate(great100 = ifelse(cases >= 100, T, F)) %>%

  head()
```

17. How many countries have average daily case totals greater than 100?

```{r}
coronavirus %>%
  group_by(country) %>% 
  summarize(average_cases = mean(cases)) %>%
  ungroup() %>%
  filter(average_cases > 100) %>%

  head()
```

18. How many countries do we have data for in the northern (lat \> 0) and southern (lat \< 0) hemispheres?

```{r}
coronavirus %>%
  mutate(hemisphere = ifelse(lat > 0, 'north', 'south')) %>%
  select(country, hemisphere) %>% 
  distinct(country, hemisphere) %>%
  count(hemisphere)
```

## Graphics

1.  Make scatter plot

```{r}
wine_data %>% 
  ggplot(aes(Color, Hue)) +
  geom_point()
```

2.  add some other parameter in color and shape make scatter plot

```{r}
wine_data %>% 
  ggplot(aes(x=Color, y=Hue, color=Alcohol, shape = Type)) +
  geom_point()
```

3.  add fit line and change theme

```{r}
wine_data %>% 
  ggplot(aes(x=Color, y=Hue, color = Type)) +
  geom_point()+
  geom_smooth() +
  theme_bw()
```

4.  make fit line linear

```{r}
wine_data %>% 
  ggplot(aes(x=Color, y=Hue, color = Type)) +
  geom_point()+
  geom_smooth(method = 'lm') +
  theme_bw()
```

5.  make boxplot of color split and color by "Type"

```{r}
wine_data %>% 
  ggplot(aes(x=Type, y=Color, fill = Type)) +
  geom_boxplot()+
  theme_classic()
```

6.  make density plot of Alcohol content and split by "Type" to make it more informative make fill transparent with smaller alpha

```{r}
wine_data %>% 
  ggplot(aes(x=Alcohol, fill = Type)) +
  geom_density(alpha=0.2)+
  theme_classic()
```

7.  another option to make density plot more readeble is to facet plot by Type, so wi add facet

```{r}
wine_data %>% 
  ggplot(aes(x=Alcohol, fill = Type, color=Type)) +
  geom_density(alpha=0.2) +
  facet_wrap(~Type) +
  theme_classic()
```

## Statistics

### T-test

1.  to test difference in means we use t test, since with t.test we can only test two groups we will silter one group out in wine example set H0: In stat test we always assume there is no difference

```{r}
wine_data %>% 
  filter(Type != "Barolo") %>% 
  t.test(Alcohol ~ Type, data = .) %>% 
  tidy()

# conclusion there is statistical sig. difference in alcohol content of two winw types
```

2.  to test more parameter we will use ANOVA

```{r}
wine_data %>% 
  aov(Alcohol ~ as.factor(Type), data = .) %>% tidy()

# conclusion at least one group is different 
```

3.  to figure out which group is diffenerent we us post hoc test `TukeyHSD()`

```{r}
wine_data %>% 
  aov(Alcohol ~ as.factor(Type), data = .) %>% TukeyHSD() %>% tidy()

# conclusion Barolo-Barbera and Grignolino-Barbera are significantly different from each other 
# but groups there is no sig. difference beteewn Grignolino-Barolo btw we could have this idea from boxplot of 
# all Type ploted together
```

### Linear Model (LM)

4.  make a linear model of two parameter from wine data set again H0: there is no correlation Alcohol and Color, these two measurement are independent

```{r}
wine_data %>% 
  ggplot(aes(Color, Alcohol)) +
  geom_point()+
  geom_smooth(method = "lm")

wine_data %>% 
  lm(Alcohol ~ Color, data = .) %>% 
  summary() %>% 
  tidy()

## conclusion there is statistically significant correlation between Alcohol and Color, this to parameters are not independent
```

5.  if we add type into calculation

```{r}
wine_data %>% 
  ggplot(aes(Color, Alcohol, color=Type)) +
  geom_point()+
  geom_smooth(method = "lm")

wine_data %>% 
  lm(Alcohol ~ Color + Type, data = .) %>% 
  summary() %>% 
  tidy()

## conclusion: this is statistically significant correlation between Alcohol and Color apply across all type
```
